# Results

```{r}
library(tidyverse)
library(ggplot2)
library(dplyr)
library(vcd)
library(nbastatR)
```

To begin our analysis, it seemed logical to first gain a sense of which teams have recorded the most total chokes, as well as comebacks over the last 4 seasons. As a result, we used a Cleveland Dot plot to display a ranking of these teams.

##Total Chokes and Comebacks

```{r}
choke_list_frame <- read.csv(file="/Users/elijahflomen/Desktop/choke_list_full_R.csv")

# Finds all instances where a home team choked
home_team_chokes <- choke_list_frame %>% filter(home_choke == 1) %>% count(home_team)
# Finds all instances where an away team choked
away_team_chokes <- choke_list_frame %>% filter(home_choke == 0) %>% count(away_team)

# Finds all instances of a home team comeback
home_team_comebacks <- choke_list_frame %>% filter(home_choke == 0) %>% count(home_team)
# Finds all instances of an away team comeback
away_team_comebacks <- choke_list_frame %>% filter(home_choke == 1) %>% count(away_team)

# Aggregates total overall chokes for a all teams
team_total_chokes <- transmute(home_team_chokes, home_team, total_chokes=home_team_chokes$n + away_team_chokes$n)
team_total_chokes <- team_total_chokes %>% rename(team=home_team) %>% arrange(team_total_chokes$total_chokes)

# Aggregates total overall comebacks for all teams
team_total_comebacks <- transmute(home_team_comebacks, home_team, total_comebacks=home_team_comebacks$n + away_team_comebacks$n)
team_total_comebacks <- team_total_comebacks %>% rename(team=home_team) %>% arrange(team)

# GRAPH: Cleveland dot plot for total chokes per team
total_chokes_plot <- ggplot(team_total_chokes, aes(reorder(team, total_chokes), total_chokes)) + geom_point(stat='identity') + coord_flip()
# GRAPH: Cleveland dot plot for total comebacks per team
total_comebacks_plot <- ggplot(team_total_comebacks, aes(reorder(team, total_comebacks), total_comebacks)) + geom_point(stat='identity') + coord_flip()

total_chokes_plot
total_comebacks_plot
```

It is interesting to note here initially the top 5 choke teams as: Memphis, Minnesota, Miami, New Orleans, Charlotte and the top 5 comeback teams as Oklahoma City, Boston, Denver, Toronto and Los Angeles. We will use these teams in a moment to further explore the possible relationships between high choke / comeback teams and other variables.

##Choke and Comeback Relationship

Further, it should be noted that being a high-comeback team and being a high-choke team are not entirely mutually exclusive. This initial data visualization showed that it is possible that a team can very well lie in both categories. A team can be simply extremely volatile (or stable) once they have secured a 90% (or fallen to a 10%) win probability, however the direction of that volatility is not consistent. Although there is no overlap within the top 5 teams in each category, it would be useful to determine if certain teams are just extremely volatile in both direction, given a sizable lead or disadvantage. 

```{r}
team_totals <- team_total_chokes %>% left_join(team_total_comebacks, by='team')
choke_combeack_scatter <- ggplot(team_totals, aes(x=total_chokes, y=total_comebacks)) + 
  geom_jitter(position = position_jitter(width=1, height=1, seed = 1)) + 
  geom_text(position = position_jitter(width=1, height=1, seed = 1), aes(label=team), hjust=0, vjust=0, size=3)
choke_combeack_scatter
```

Some interesting findings from this scatter plot is that there are indeed some teams with notable volatility (and stability) in their game play. For instance, Charlotte, a team we already identified to be within the top 5 choke teams also has a sizable number of comebacks. Further, Denver, a team that ranked in the top 5 comeback teams, was also within the top half of choke-prone teams. As a result, these two teams stand out in that for both of them, they stand out, relative to the rest of the league, in that their games often result in improbable endings, for better and for worse. Therefore, it would be of interest to investigate further aspects of these teams that cause them to stand out as improbable ending prone teams. Additionally, it was not surprising to notice that the number 1 team in both chokes and comebacks, MEM and OKC respectively, both ranked among the lowest in their respective opposite category. 

##Home v. Visitor impact on Choke v. Comeback

Given these top teams in both choke and comeback categories, we wanted to investigate whether or not being the home vs. visitor team was a significant determinant of the outcome of the game. In the sports community, it is often known that home teams typically have a slight advantage, however, we wanted to investigate the nuances of this advantage at the choke/comeback level. For all teams across the league, we have produced mosaic plots to get a visual representation of this potential relationship. We split the league in half, by alphabetical order just to have the graph not be overly crowded and to view the relationship more clearly. 

```{r}

home_team_chokes$loc <- "h"
home_team_chokes$result <- "choke"
home_team_comebacks$loc <- "h"
home_team_comebacks$result <- "comeback"
home_team_chokes <- home_team_chokes %>% rename(team=home_team)
home_team_comebacks <- home_team_comebacks %>% rename(team=home_team)


away_team_chokes$loc <- "v"
away_team_chokes$result <- "choke"
away_team_comebacks$loc <- "v"
away_team_comebacks$result <- "comeback"
away_team_chokes <- away_team_chokes %>% rename(team=away_team)
away_team_comebacks <- away_team_comebacks %>% rename(team=away_team)


joined_team_data <- rbind(home_team_chokes, home_team_comebacks, away_team_chokes, away_team_comebacks)
longer_joined_team_data <- joined_team_data %>% uncount(n)

longer_joined_team_data_1 <- longer_joined_team_data %>%
  filter(str_detect(team,'ATL|BKN|BOS|CHA|CHI|CLE|DAL|DEN|DET|GSW|HOU|IND|LAC|LAL|MEM'))

longer_joined_team_data_2 <- longer_joined_team_data %>%
  filter(str_detect(team,'MIA|MIL|MIN|NOP|NYK|OKC|ORL|PHI|PHX|POR|SAC|TOR|UTA|WAS|SAS'))

mosaic(result~team+loc, direction=c('v', 'v', 'h'), 
       data=longer_joined_team_data_1)

mosaic(result~team+loc, direction=c('v', 'v', 'h'), 
       data=longer_joined_team_data_2)
```

As we can see from the following two mosaic plots, some teams exhibit significant differences between their number of chokes and comebacks when they are home vs. when they are a visitor. What was most surprising from this visualization was that a vast majority of teams have a much higher comeback rate when they are a visitor. While most teams have this trend, teams such as DER and TOR show a strong positive relationship between them experiencing a comeback and being a visitor team. Perhaps a logical explanation for this finding is that home teams can do a better job of starting the game off strong, using the crowd and familiar environment to their advantage. As a result, many of the instances of comebacks for a team occur when they are a visitor, simply due to the fact that a home team has a better chance of gaining an early lead. On the flip side, for these same teams, they had much more chokes as home teams than as visitors. Along the same logic, this can likely be interpreted as home teams having the advantage of starting off a game very strong, but as a result, lends itself to the possibility of a choke in the latter part of the game. This finding was extremely curious to us since it caused us to question the true nature of having home-court advantage. However, given that this dataset is only considering the games that had a choke/comeback ending, this seemingly contradictory relationship is actually quite logical.  

##Reasons for Choke, split by Home v. Away

Now, given that there is some explanatory connection between being home vs. away and the outcome of a team’s game, we wanted to leverage the play-by-play data to see the most frequent events in a game that resulted in a choke. This analysis would enable us to assign a single cause to a choke instance and then examine holistically which of these causes are most common, split by home chokes vs. away chokes. We decided that a fair metric to determine the single cause for a choke would be to identify the event with the maximum standardized frequency. As mentioned earlier, the main 3 mistakes that a team makes which leads them to a choke include: misses, fouls and turnovers. However, it would be naive to simply count the frequencies of these events in a choke and then assign the event with the largest count to be the explanatory mistake. This is because in a game, naturally some of these events will occur more often than others, regardless of the final game outcome. To fix this, we looked at the ratio of the counts of each mistake, relative to the total league averages. We obtained these league averages from this website. (https://www.basketball-reference.com/leagues/NBA_stats_per_game.html). Since we standardized the frequencies of each mistake to the average, we can then identify the mistakes that occurred at the largest multiple, against the league average. As an example, if the number of fouls in a game is around 10, if a team commits 20 fouls, it is likely that since they committed 2x of the league average, their abundance of fouls was the main contributor for their choke. We used a bar chart to display the distribution of games that has misses, fouls or turnovers as their main explanatory variable.


```{r}
pbp_frame <- read.csv(file="/Users/elijahflomen/Desktop/pbp_data_full_R.csv")
home_choke_pbp <- pbp_frame %>% group_by(game_id) %>% filter(max(visitor_pts) > max(home_pts))
home_choke_pbp <- home_choke_pbp %>% mutate(home_mistake = ifelse(grepl("Turnover", home_desc), 'turnover', 
                                                           ifelse(grepl("FOUL", home_desc), 'foul',
                                                           ifelse(grepl("MISS", home_desc), 'miss', '')))) 


# Repeat the same process for the away team chokes play-by-play data
away_choke_pbp <- pbp_frame %>% group_by(game_id) %>% filter(max(home_pts) > max(visitor_pts))
away_choke_pbp <- away_choke_pbp %>% mutate(away_mistake = ifelse(grepl("Turnover", visitor_desc), 'turnover', 
                                                           ifelse(grepl("FOUL", visitor_desc), 'foul',
                                                           ifelse(grepl("MISS", visitor_desc), 'miss', '')))) 


# Here, we want to look to classifiy each choke by a single cause, to determine the leading causes for a team to lose
home_choke_reasons <- home_choke_pbp %>% group_by(game_id) %>% count(home_mistake) %>% na_if("") %>% na.omit
away_choke_reasons <- away_choke_pbp %>% group_by(game_id) %>% count(away_mistake) %>% na_if("") %>% na.omit

# We do this by looking at the count of each mistake for a team that choked per game. 
# We then want to standardize the frequency of these mistakes by dividing them by the league averages
# This method allows us to identify the mistake that was most frequent in a choke
home_choke_reasons_frac <- home_choke_reasons %>% mutate(frac = ifelse(home_mistake == 'turnover', n/14.3,
                                                                ifelse(home_mistake == 'miss', n/48,
                                                                ifelse(home_mistake == 'foul', n/19, 0))))

away_choke_reasons_frac <- away_choke_reasons %>% mutate(frac = ifelse(away_mistake == 'turnover', n/14.3,
                                                               ifelse(away_mistake == 'miss', n/48,
                                                               ifelse(away_mistake == 'foul', n/19, 0))))

home_choke_reason_max <- home_choke_reasons_frac %>% group_by(game_id) %>% filter(frac==max(frac))
away_choke_reason_max <- away_choke_reasons_frac %>% group_by(game_id) %>% filter(frac==max(frac))

# GRAPH 3: Leading Causes for a home-team choke
ggplot(home_choke_reason_max) + geom_bar(aes(home_mistake))

# GRAPH 4: Leading Causes for an away-team choke
ggplot(away_choke_reason_max) + geom_bar(aes(away_mistake))
```

As we can see here, the clear leading cause for both home and away chokes are turnovers. Given that turnovers are the clear leading causes for chokes over the last 4 seasons, it would be of interest to investigate the turnover rates of specific players in the NBA, which will follow in our analysis. An additional interesting finding is that, for home chokes, misses are more frequent than fouls, but the opposite is true for away chokes. Therefore, fouls had a larger presence than misses for team to choke, when they are playing as a visitor. 

##Player Stat Heat Maps

Now that we have identified the teams with the highest chokes / comebacks, as well as the most frequent causes for these outcomes, it would be of interest to dive into some of the player-specific statistics. Given that fouls, misses, turnovers are all the cause of players’ specific actions, it would be of good use to visualize these, as well as other individual player-level statistics. This investigation would allow us to determine if, for high-choke teams, the main reasons for their unfortunate endings are the result of one specific player, or rather the team as a whole performing poorly in a specific category. The highest-choke team in our data set was found to be The Memphis Grizzlies, therefore, it would be interesting to further explore their individual players’ statistics. Symmetrically, we also wanted to visualize the individual player-level statistics of the highest comeback team, the Oklahoma City Thunder.

A natural way to express player-level data is in the form of a heat map, showing the average ranking of that player in that statistic, over the course of their career. Given that players are often moved around teams, there is no perfect system to determine their statistics, filtered down to only the games when they played for a specific team, however, we simply decided to look at the individual player statistics who were most frequently on that team during their chokes/comebacks. This seemed like the most logical approach to solve the issue of inconsistent rosters, since we are looking at the player career stats of individuals who played on frequently choke / comeback teams.


```{r}

mem_starter_stats <- 
  players_careers(players=c("Ja Morant", "Kyle Anderson", "Jaren Jackson Jr.", "Jonas Valanciunas", "Marc Gasol"),
                  modes=c("Totals"), 
                  assign_to_environment = FALSE,
                  add_mode_names = FALSE, return_message = FALSE) %>% filter(nameTable=="SeasonRankingsRegularSeason")

morant_stats <- mem_starter_stats %>% filter(namePlayer=="Ja Morant") %>% select(dataTable)
morant_stats <- morant_stats[[1]][[1]] %>% select(fg3mRank, fgmRank, pctFGRank, ptsRank, tovRank, trebRank)
morant_stats$player <- "Ja Morant"
morant_longer <- morant_stats[complete.cases(morant_stats),] %>% pivot_longer(!player, names_to = 'stat', values_to = 'value') 
morant_longer <- morant_longer %>% group_by(stat) %>% summarize(mean(value, na.rm=TRUE))
morant_longer$player <- "Ja Morant"

anderson_stats <- mem_starter_stats %>% filter(namePlayer=="Kyle Anderson") %>% select(dataTable)
anderson_stats <- anderson_stats[[1]][[1]] %>% select(fg3mRank, fgmRank, pctFGRank, ptsRank, tovRank, trebRank)
anderson_stats$player <- "Kyle Anderson"
anderson_longer <- anderson_stats %>% pivot_longer(!player, names_to = 'stat', values_to = 'value')
anderson_longer <- anderson_longer %>% group_by(stat) %>% summarize(mean(value, na.rm=TRUE))
anderson_longer$player <- "Kyle Anderson"

jackson_stats <- mem_starter_stats %>% filter(namePlayer=="Jaren Jackson Jr.") %>% select(dataTable)
jackson_stats <- jackson_stats[[1]][[1]] %>% select(fg3mRank, fgmRank, pctFGRank, ptsRank, tovRank, trebRank)
jackson_stats$player <- "Jaren Jackson Jr."
jackson_longer <- jackson_stats %>% pivot_longer(!player, names_to = 'stat', values_to = 'value')
jackson_longer <- jackson_longer %>% group_by(stat) %>% summarize(mean(value, na.rm=TRUE))
jackson_longer$player <- "Jaren Jackson Jr."

valanciunas_stats <- mem_starter_stats %>% filter(namePlayer=="Jonas Valanciunas") %>% select(dataTable)
valanciunas_stats <- valanciunas_stats[[1]][[1]] %>% select(fg3mRank, fgmRank, pctFGRank, ptsRank, tovRank, trebRank)
valanciunas_stats$player <- "Jonas Valanciunas"
valanciunas_longer <- valanciunas_stats %>% pivot_longer(!player, names_to = 'stat', values_to = 'value')
valanciunas_longer <- valanciunas_longer %>% group_by(stat) %>% summarize(mean(value, na.rm=TRUE))
valanciunas_longer$player <- "Jonas Valanciunas"

gasol_stats <- mem_starter_stats %>% filter(namePlayer=="Marc Gasol") %>% select(dataTable)
gasol_stats <- gasol_stats[[1]][[1]] %>% select(fg3mRank, fgmRank, pctFGRank, ptsRank, tovRank, trebRank)
gasol_stats$player <- "Marc Gasol"
gasol_longer <- gasol_stats %>% pivot_longer(!player, names_to = 'stat', values_to = 'value')
gasol_longer <- gasol_longer %>% group_by(stat) %>% summarize(mean(value, na.rm=TRUE))
gasol_longer$player <- "Marc Gasol"

mem_total_heat <- rbind(morant_longer, jackson_longer, anderson_longer, valanciunas_longer, gasol_longer) 
colnames(mem_total_heat)[2] <- "rank"

ggplot(data=mem_total_heat, aes(player, stat, fill=rank)) + geom_tile()

```





